Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-03 paper/paper.tex
misc revision

  \item \textbf{Refunctionalization}~\cite{DANVY2009534,Danvy:2006:RW:2171265.2171268}:
    We apply refunctionalization to the disentangled AAM, which
    %as the left inverse of Reynolds' defunctionalization \cite{Reynold:72,Reynolds:HOSC98-revisited} 
    replaces the first-order data types representing continuations with
    higher-order functions, and associates dispatching logic with proper higher-order functions. 
    We obtain an abstract interpreter written in continuation-passing style with an
    additional layer of continuations due to nondeterminism.
%
    For clarity, we first present the refunctionalized AAM without caching. 
    We then adopt a caching algorithm \cite{darais2017abstracting} to guarantee the termination 
    of abstract interpretation.
    In the end of this section, we review the pushdown control-flow analysis and examine how 
    computable and precise call/return matching is obtained through these transformations
    (Section~\ref{sec:refunc}).

@@

\begin{lstlisting}
  def drive(nds: NDState, seen: Set[State]): Set[State] = {
    val s = nds.toState;  val new_seen = if (seen.contains(s)) seen else seen + s
    step(nds) match {
      case None => new_seen
      case Some(nnds) => drive(nnds, new_seen)
    }
  }
\end{lstlisting}

The @drive@ function is also changed: there is no work-list anymore, as all the potentially 
unexplored states are embedded into the meta-continuation.
The termination of the analysis occurs when the value @None@ is returned from calling @step@.
Otherwise, we extract the next state @nnds@ and recur on function @drive@.
The set of explored states is preserved as before.

@@

\begin{itemize}
  \item The components of the configurations are lifted as arguments of the evaluation
    function.
  \item Since our final target is an abstract definitional interpreter (though still
    nondeterministic), starting from this section our abstract semantic artifact
    returns a set of final values instead of collected configurations.
  \item Configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible concrete reachable paths and values.
  \item We simplify the continuations with an nondeterministic operator @nd@ such 
    that the function @aeval@ only uses one layer of continuation.
\end{itemize}

@@

As previously identified, there are two layers of continuations: one for the normal stack 
of the analyzed language, and the other for the nondeterministic choices of closures.
This corresponds to the extended continuation-passing style.
But in our defining language (Scala), the delimited control operators rely on the continuation-passing 
transformation which is only one layer.

@@

Now we have arrived at the end of this series of transformations. 
The inverse transformation of the last step is basically CPS transforming the function @aeval@ and @nd@.
We may first explicitly expose the continuation at the level of @aeval@, and then convert @nd@ 
to continuation-passing style. Besides, these continuations have to be manifestly embedded
at the call-site of @nd@, instead of implicitly provided by the @shift@ operator.

To conclude, starting from a small-step abstract abstract machine, eventually we obtain a 
big-step definitional interpreter with pushdown control-flow, written in direct-style.

>>>>

  \item \textbf{Refunctionalization}~\cite{DANVY2009534,Danvy:2006:RW:2171265.2171268}:
    We apply refunctionalization to the disentangled AAM, which
    %as the left inverse of Reynolds' defunctionalization \cite{Reynold:72,Reynolds:HOSC98-revisited} 
    replaces the first-order data types representing continuations with
    higher-order functions, and associates dispatching logic with proper higher-order functions. 
    We obtain an abstract interpreter written in continuation-passing style with an
    additional layer of continuations due to nondeterminism.

    For clarity, we first present a vanilla version of refunctionalized AAM which simply
    converts the stack structures to higher-order functions but keeps other parts unchanged.
    We then adopt a caching algorithm \cite{darais2017abstracting} to guarantee the termination 
    of abstract interpretation.
    In the end of this section, we review the pushdown control-flow analysis and examine how 
    computable and precise call/return matching is obtained through these transformations
    (Section~\ref{sec:refunc}).

@@

\begin{lstlisting}
  def drive(nds: NDState, seen: Set[State]): Set[State] = {
    val s = nds.toState;  val new_seen = if (seen.contains(s)) seen else seen + s
    step(nds) match {
      case None => new_seen;  case Some(nnds) => drive(nnds, new_seen)
    } }
\end{lstlisting}

The @drive@ function is also changed: there is no work-list anymore, as all the potentially 
unexplored states are embedded into the continuations.
The termination of the analysis occurs when the value @None@ is returned from calling @step@.
Otherwise, we extract the next state @nnds@ and recur on function @drive@.
The set of explored states is preserved as before.

@@

\begin{itemize}
  \item The components of the configurations are lifted as arguments of the evaluation
    function.
  \item The refunctionalized AAM collects all possible configurations, but our final target is an 
    abstract definitional interpreter that cares about final evaluation results.
    So starting from this section our abstract semantic artifact returns a set of final 
    values instead of collected configurations. 
  \item Configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible concrete reachable paths and values.
  \item We simplify the continuations with an nondeterministic operator @nd@ such 
    that the function @aeval@ only uses one layer of continuation.
\end{itemize}

@@

As previously identified, there are two layers of continuations: one for the normal stack 
structure of the analyzed language, and the other for the nondeterministic choices of closures.
This corresponds to the extended continuation-passing style.
But in our defining language (Scala), the delimited control operators rely on the continuation-passing 
transformation which is only one layer.

@@

Now we have arrived at the end of this series of transformations. 
The inverse transformation of the last step is basically CPS transforming the function @aeval@ and @nd@.
We may first explicitly expose the continuation at the level of @aeval@, and then convert @nd@ 
to extended continuation-passing style. Besides, these continuations must be manifestly embedded
at the call-site of @nd@, instead of implicitly provided by the @shift@ operator.

To conclude, starting from a small-step abstract abstract machine, eventually we obtain a 
big-step abstract definitional interpreter with pushdown control-flow, written in direct-style.
